Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)

### 2015-04-27 WSSSPE2_report/WSSSPE2_review.tex
Section 4: Added references for all talks.

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\subsection{Kaitlin Thaney, Designing for Truth, Scale, and Sustainability}
\label{keynote1}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

@@

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\subsection{Neil Chue Hong, We are the 92$\%$}
\label{keynote2}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{comment}
In a recent survey of UK research-intensive universities, 92\% of researchers
said they used research software and 68\% said their research would be impossible
without software. Yet 71\% have had no formal software training, and few are
ready to apply many of the things we take for granted such as testing or
virtualization. WSSSPE represents the pinnacle of what we understand to be the
best practice around scientific software in our community. My talk will
challenge the workshop participants to come up with ways of taking this best
practice to those 92\% of researchers in a way that will lead to maximum benefit
to the scientific community.
 \end{comment}

@@

Neil Chue Hong is the Director of the Software Sustainability Institute (SSI) in the 
United Kingdom. The SSI was founded to support the UK's research software 
community by cultivating better, more sustainable research software to enable
world-class research. Chue Hong's keynote\footnote{Slides available at:
http://dx.doi.org/10.6084/m9.figshare.1243288} started by making the point of
that the use of -- and reliance on -- software is pervasive in all areas of
world-leading research, showing examples from disciplines as diverse as
humanities and high-energy physics, and quoting Kersten Kleese Van Dam of the
Pacific Northwestern National Laboratory via a petition
campaign\footnote{https://www.change.org/p/everyone-in-the-research-community-we-must-accept-that-software-is-fundamental-to-research-or-we-will-lose-our-ability-to-make-groundbreaking-discoveries}
at change.org, ``\emph{Today there are very few science areas left which do not
rely on IT and thus software for the majority of their research work. More
importantly key scientific advances in experimental and observational science
would have been impossible without better software.}'' He also cited Daniel
Katz, Software Infrastructure for Sustained Innovation (SI2) Program Director of
the NSF, ``\emph{Scientific discovery and innovation are advancing along
fundamentally new pathways opened by development of increasingly sophisticated
software. Software is an integral enabler of computation, experiment and theory,
and directly responsible for increased scientific productivity and enhancement
of researchers' capabilities.}''

@@

\begin{enumerate}
%\item Greg Wilson: Close Enough for Scientific Work

\item \textbf{Colin Venters: The Nebuchadnezzar Effect: Dreaming of Sustainable Software
through Sustainable Software Architectures.}
Venters proposed that sustainable software is a composite, first-class,
non-functional requirement (NFR) that is at a minimum a measure of a system's
maintainability and extensibility, but may also include other NFRs such as
efficiency (e.g., energy, cost), interoperability, portability, reusability,
scalability, and usability. To achieve technically sustainable software, Venters
suggested that software architectures are fundamental as they are the primary
carrier of system NFRs, i.e., pre-system understanding; and influence how
developers are able to understand, analyze, extend, test, and maintain a
software system, i.e., post-deployment system understanding. In addition,
Venters highlighted that sustainability of software architectures needs to be
addressed to endure different types of change and evolution in order to mitigate
architectural drift, erosion, and knowledge vaporization.

@@

\item \textbf{Marlon Pierce: Patching It Up, Pulling It Forward.}
Pierce discussed how open open source is. Open software needs a diverse, openly
governed community behind it, just as it needs open licensing and a public code
repository. To probe the level of governance within open source projects, he and
his co-authors (Suresh Marru and Chris Mattmann) suggested a contest to encourage
individual developers to submit patches and pull request to projects that are
important to them. This simple mechanism will expose several governance
mechanisms, such as how easy it is for independent developers to communicate
with project leadership, how projects accept and properly license third party
contributions, and how projects make decisions such as granting source tree
write access.

\item \textbf{John Peterson: Continuous Integration for Concurrent MOOSE Framework and
Application Development on GitHub.} 
%Affiliation: Idaho National Laboratory
Peterson from the Idaho National Laboratory reported that in March 2014, the
MOOSE framework was released under an open source license on GitHub,
significantly expanding and diversifying the pool of current active and
potential future contributors on the project. The MOOSE team employs an
extensive continuous integration test suite to ensure that both the framework
and the applications based on the framework are verified before any code changes
are accepted into the repository. They use a combination of built-in Git
features such as branching and submodules, GitHub API integration capabilities,
and in-house developed testing software to perform this verification and update
the dependent applications in a relatively seamless manner for users.

\item \textbf{Abani Patra: Execute it.}
%Univ at Buffalo, SUNY
Patra discussed the value of
an available and easily accessible platform for executing scientific software,
e.g., HUBzero, to access XSEDE or other computing resources.
Such a platform for executing benchmark problems (even at small
scale) allows the developer community access to a reference implementation and
provides an easy way to train the larger user community. A second idea of this
presentation was that for true usability, much more attention and support needs to be given to the
integrated use of simulation tools inside complex workflows.
%needs supporting the user community workflow.

@@

\item \textbf{Daniel Katz: Implementing Transitive Credit with JSON-LD.}
Science and engineering research increasingly relies on activities that
facilitate research but are not rewarded or recognized, such as: data
sharing; developing common data resources, software and methodologies; and
annotating data and publications. To promote and advance these activities, we
must develop mechanisms for assigning credit, facilitate the appropriate
attribution of research outcomes, devise incentives for activities that
facilitate research, and allocate funds to maximize return on investment. Katz
discussed the issue of assigning credit for both direct and indirect
contributions by using JSON-LD to implement a prototype transitive credit
system~\cite{Katz_transitive_credit_poster}.

\item \textbf{Samin Ishtiaq: Daemons, Notifications and Sustaining Software.}
%  Microsoft Research Cambridge
The reproduction and replication of novel results has become a major issue in
computer science, systems biology, and other computational disciplines. These
include both the inability to re-implement novel algorithms and approaches, and
lack of an agreement on how and what to benchmark these algorithms on.
Ishtiaq from Microsoft Research Cambridge pointed out these problems and 
made several suggestions to address them.

\item \textbf{James Howison: Retract all Bit-Rotten Publications.}
Howison sought to provoke discussion by proposing that papers whose workflows are
not kept current with the changing software ecosystem should be automatically
retracted. This would create an incentive for authors to keep their
software current and usable, rather than the current situation in which every
potential user has to do this individually. A softer version of the proposal
would identify papers whose software workflow has become bit-rotten and allow
others to keep the code up to date, either adding them as new authors of the
paper or providing credit in some other form for their academic service.

@@

\item \textbf{Robert Downs: Community Recommendations for Improving Sustainable
Scientific Software Practices.}
Robert Downs, of the Columbia University Center for International Earth
Science Information Network (CIESIN), 
%presented Community Recommendations for
%Improving Sustainable Scientific Software Practices (2014), and 
described a
focus group study conducted with %W. Christopher Lenhardt, Erin Robinson, Ethan
% Davis, and Nicholas Weber of 
the Science Software Cluster (SSC) of the
Federation of Earth Science Information Partners (ESIP). For the study, almost
300 attendees of the 2014 Summer ESIP Meeting were invited to participate in
simultaneous roundtable discussions on sustainability of science software. Over
two-thirds of the roundtable focus groups responded to a semi-structured survey
that contained three sets of questions eliciting recommendations for near-term
actions of the community to improve sustainable software practices. Initial
analysis of the
participants' responses to the questionnaire revealed several suggestions, which
included improving community engagement and collaborative activities, increasing
understanding and awareness, and creating incentives to motivate sustainable
science software practices.
%Based on the recommendations
%offered by the focus group respondents for near-term actions, 
The ESIP SSC plans to engage the community %to achieve progress in the near future 
in the recommended activities for improving sustainable scientific
software practices.

@@

\item \textbf{Carl Boettiger: rOpenSci: Building Sustainable Software by Fostering a
Diverse Community.}
Boettiger described how the rOpenSci project has been successful by focusing not
just on building software but also on building a community of researchers who
learn and adopt their approaches to reproducible research and sustainable
software practice. Through outreach, mentoring, workshops, and hackathons, they
have not only reached new users, but also turned users into co-developers of
robust software and good practices to support data science research across a
growing set of disciplines.

\item \textbf{Jakob Blomer: The Need for a Versioned Data Analysis Environment}
%
Large-scale scientific endeavors, such as the discovery of the Higgs
boson at the Large Hadron Collider (LHC), often rely on complex software
stacks.  Maintaining thousands of dependencies of software
libraries and operating system versions has shown that despite
source code availability, the setup and the validation of a minimal
usable analysis environment can easily become prohibitively expensive.
%There is a substantial gap between merely having access to open source
%software and the ability to create a data analysis runtime environment.
 In high-energy physics, a
special purpose, open-source, versioning and snapshotting file system
(CernVM-FS) that is used to capture and distribute entire software stacks,
providing instant access to ready-to-run data analysis environments,
proved to be very useful.
% for the global scientific community as well as for
%interested citizens, for instance through the CERN Open Data Portal
%(http://opendata.cern.ch). \todo{revise}

\item \textbf{Alice Allen: Find it! Cite it!}
%
The Astrophysics Source Code Library (ASCL) is an online registry of
scientist-written software used in astronomy research. Their primary interest is
rendering research more transparent by making this software more discoverable for
examination. The ASCL is treated as a publication by an indexing resource for
astronomy, the Astrophysics Data System (ADS). ADS tracks citations to what it
indexes, including citations to ASCL entries. Increasing rewards for writing
software, whether through citation, transitive credit or other methods, gives
software authors a powerful reason to take the time to build sustainability into
their software and is an excellent way to drive community change.

\begin{comment}
The ASCL actively looks for opportunities to use change management strategies to
move astronomy to a more transparent discipline; these strategies can be used in
other disciplines to further change in them.
\end{comment}
\end{enumerate}

>>>>

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\subsection{Kaitlin Thaney, Designing for Truth, Scale, and Sustainability~\cite{Thaney_slides}}
\label{keynote1}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

@@

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\subsection{Neil Chue Hong, We are the 92$\%$~\cite{Hong_slides}}
\label{keynote2}
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\begin{comment}
In a recent survey of UK research-intensive universities, 92\% of researchers
said they used research software and 68\% said their research would be impossible
without software. Yet 71\% have had no formal software training, and few are
ready to apply many of the things we take for granted such as testing or
virtualization. WSSSPE represents the pinnacle of what we understand to be the
best practice around scientific software in our community. My talk will
challenge the workshop participants to come up with ways of taking this best
practice to those 92\% of researchers in a way that will lead to maximum benefit
to the scientific community.
 \end{comment}

@@

Neil Chue Hong is the Director of the Software Sustainability Institute (SSI) in the 
United Kingdom. The SSI was founded to support the UK's research software 
community by cultivating better, more sustainable research software to enable
world-class research. Chue Hong's keynote started by making the point of
that the use of -- and reliance on -- software is pervasive in all areas of
world-leading research, showing examples from disciplines as diverse as
humanities and high-energy physics, and quoting Kersten Kleese Van Dam of the
Pacific Northwestern National Laboratory via a petition
campaign\footnote{https://www.change.org/p/everyone-in-the-research-community-we-must-accept-that-software-is-fundamental-to-research-or-we-will-lose-our-ability-to-make-groundbreaking-discoveries}
at change.org, ``\emph{Today there are very few science areas left which do not
rely on IT and thus software for the majority of their research work. More
importantly key scientific advances in experimental and observational science
would have been impossible without better software.}'' He also cited Daniel
Katz, Software Infrastructure for Sustained Innovation (SI2) Program Director of
the NSF, ``\emph{Scientific discovery and innovation are advancing along
fundamentally new pathways opened by development of increasingly sophisticated
software. Software is an integral enabler of computation, experiment and theory,
and directly responsible for increased scientific productivity and enhancement
of researchers' capabilities.}''

@@

%\item Greg Wilson: Close Enough for Scientific Work

\subsection{Colin Venters: The Nebuchadnezzar Effect: Dreaming of Sustainable Software
through Sustainable Software Architectures~\cite{Venters_poster}}
Venters proposed that sustainable software is a composite, first-class,
non-functional requirement (NFR) that is at a minimum a measure of a system's
maintainability and extensibility, but may also include other NFRs such as
efficiency (e.g., energy, cost), interoperability, portability, reusability,
scalability, and usability. To achieve technically sustainable software, Venters
suggested that software architectures are fundamental as they are the primary
carrier of system NFRs, i.e., pre-system understanding; and influence how
developers are able to understand, analyze, extend, test, and maintain a
software system, i.e., post-deployment system understanding. In addition,
Venters highlighted that sustainability of software architectures needs to be
addressed to endure different types of change and evolution in order to mitigate
architectural drift, erosion, and knowledge vaporization.

@@

\subsection{Marlon Pierce: Patching It Up, Pulling It Forward~\cite{Pierce_poster}}
Pierce discussed how open open source is. Open software needs a diverse, openly
governed community behind it, just as it needs open licensing and a public code
repository. To probe the level of governance within open source projects, he and
his co-authors (Suresh Marru and Chris Mattmann) suggested a contest to encourage
individual developers to submit patches and pull request to projects that are
important to them. This simple mechanism will expose several governance
mechanisms, such as how easy it is for independent developers to communicate
with project leadership, how projects accept and properly license third party
contributions, and how projects make decisions such as granting source tree
write access.

\subsection{John Peterson: Continuous Integration for Concurrent MOOSE Framework and
Application Development on GitHub~\cite{Peterson_poster}} 
%Affiliation: Idaho National Laboratory
Peterson from the Idaho National Laboratory reported that in March 2014, the
MOOSE framework was released under an open source license on GitHub,
significantly expanding and diversifying the pool of current active and
potential future contributors on the project. The MOOSE team employs an
extensive continuous integration test suite to ensure that both the framework
and the applications based on the framework are verified before any code changes
are accepted into the repository. They use a combination of built-in Git
features such as branching and submodules, GitHub API integration capabilities,
and in-house developed testing software to perform this verification and update
the dependent applications in a relatively seamless manner for users.

\subsection{Abani Patra: Execute it~\cite{Patra_poster}}
%Univ at Buffalo, SUNY
Patra discussed the value of
an available and easily accessible platform for executing scientific software,
e.g., HUBzero, to access XSEDE or other computing resources.
Such a platform for executing benchmark problems (even at small
scale) allows the developer community access to a reference implementation and
provides an easy way to train the larger user community. A second idea of this
presentation was that for true usability, much more attention and support needs to be given to the
integrated use of simulation tools inside complex workflows.
%needs supporting the user community workflow.

@@

\subsection{Daniel Katz: Implementing Transitive Credit with JSON-LD~\cite{Katz_transitive_credit_poster}}
Science and engineering research increasingly relies on activities that
facilitate research but are not rewarded or recognized, such as: data
sharing; developing common data resources, software and methodologies; and
annotating data and publications. To promote and advance these activities, we
must develop mechanisms for assigning credit, facilitate the appropriate
attribution of research outcomes, devise incentives for activities that
facilitate research, and allocate funds to maximize return on investment. Katz
discussed the issue of assigning credit for both direct and indirect
contributions by using JSON-LD to implement a prototype transitive credit
system.

\subsection{Samin Ishtiaq: Daemons, Notifications and Sustaining Software~\cite{Ishtiaq_poster}}
%  Microsoft Research Cambridge
The reproduction and replication of novel results has become a major issue in
computer science, systems biology, and other computational disciplines. These
include both the inability to re-implement novel algorithms and approaches, and
lack of an agreement on how and what to benchmark these algorithms on.
Ishtiaq from Microsoft Research Cambridge pointed out these problems and 
made several suggestions to address them.

\subsection{James Howison: Retract all Bit-Rotten Publications~\cite{Howison_poster}}
Howison sought to provoke discussion by proposing that papers whose workflows are
not kept current with the changing software ecosystem should be automatically
retracted. This would create an incentive for authors to keep their
software current and usable, rather than the current situation in which every
potential user has to do this individually. A softer version of the proposal
would identify papers whose software workflow has become bit-rotten and allow
others to keep the code up to date, either adding them as new authors of the
paper or providing credit in some other form for their academic service.

@@

\subsection{Robert Downs: Community Recommendations for Improving Sustainable
Scientific Software Practices~\cite{Downs_poster}}
Robert Downs, of the Columbia University Center for International Earth
Science Information Network (CIESIN), 
%presented Community Recommendations for
%Improving Sustainable Scientific Software Practices (2014), and 
described a
focus group study conducted with %W. Christopher Lenhardt, Erin Robinson, Ethan
% Davis, and Nicholas Weber of 
the Science Software Cluster (SSC) of the
Federation of Earth Science Information Partners (ESIP). For the study, almost
300 attendees of the 2014 Summer ESIP Meeting were invited to participate in
simultaneous roundtable discussions on sustainability of science software. Over
two-thirds of the roundtable focus groups responded to a semi-structured survey
that contained three sets of questions eliciting recommendations for near-term
actions of the community to improve sustainable software practices. Initial
analysis of the
participants' responses to the questionnaire revealed several suggestions, which
included improving community engagement and collaborative activities, increasing
understanding and awareness, and creating incentives to motivate sustainable
science software practices.
%Based on the recommendations
%offered by the focus group respondents for near-term actions, 
The ESIP SSC plans to engage the community %to achieve progress in the near future 
in the recommended activities for improving sustainable scientific
software practices.

@@

\subsection{Carl Boettiger: rOpenSci: Building Sustainable Software by Fostering a
Diverse Community~\cite{Boettiger_poster}}
Boettiger described how the rOpenSci project has been successful by focusing not
just on building software but also on building a community of researchers who
learn and adopt their approaches to reproducible research and sustainable
software practice. Through outreach, mentoring, workshops, and hackathons, they
have not only reached new users, but also turned users into co-developers of
robust software and good practices to support data science research across a
growing set of disciplines.

\subsection{Jakob Blomer: The Need for a Versioned Data Analysis Environment~\cite{Blomer_poster}}
%
Large-scale scientific endeavors, such as the discovery of the Higgs
boson at the Large Hadron Collider (LHC), often rely on complex software
stacks.  Maintaining thousands of dependencies of software
libraries and operating system versions has shown that despite
source code availability, the setup and the validation of a minimal
usable analysis environment can easily become prohibitively expensive.
%There is a substantial gap between merely having access to open source
%software and the ability to create a data analysis runtime environment.
 In high-energy physics, a
special purpose, open-source, versioning and snapshotting file system
(CernVM-FS) that is used to capture and distribute entire software stacks,
providing instant access to ready-to-run data analysis environments,
proved to be very useful.
% for the global scientific community as well as for
%interested citizens, for instance through the CERN Open Data Portal
%(http://opendata.cern.ch). \todo{revise}

\subsection{Alice Allen: Find it! Cite it!~\cite{Allen_poster}}
%
The Astrophysics Source Code Library (ASCL) is an online registry of
scientist-written software used in astronomy research. Their primary interest is
rendering research more transparent by making this software more discoverable for
examination. The ASCL is treated as a publication by an indexing resource for
astronomy, the Astrophysics Data System (ADS). ADS tracks citations to what it
indexes, including citations to ASCL entries. Increasing rewards for writing
software, whether through citation, transitive credit or other methods, gives
software authors a powerful reason to take the time to build sustainability into
their software and is an excellent way to drive community change.

\begin{comment}
The ASCL actively looks for opportunities to use change management strategies to
move astronomy to a more transparent discipline; these strategies can be used in
other disciplines to further change in them.
\end{comment}
